Supervised Topic Modeling in Clinical Text

### 2011-04-25 paper/main.tex
Updated generative model and minor fixes

\begin{figure}[h]%tbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{model_1}
   \includegraphics[width=3in]{model_2}  
   \caption{adapted sLDA model}
   \label{fig:example}
\end{figure}

Given a fixed set of parameters: $\beta_{1:K}$, the K topics, $\alpha$,
the parameter for the document specific dirichlet distribution, and
$\eta$, the response variable parameter, the generative process for
documents and responses is as follows: 
\begin{enumerate}
\item Draw topic proportions $\theta\mid\alpha\sim Dir\left(\alpha\right)$ 
\item For each word:

\begin{enumerate}
\item Draw topic assignment $z_{n}\mid\theta\sim Mult\left(\theta\right)$ 
\item Draw word $w_{n}\mid z_{n},\beta_{1:K}\sim Mult\left(\beta_{z_{n}}\right)$ 
\end{enumerate}
\item For each ICD-9 code from the root of the hierarchy and recursively
descending the tree:

\begin{enumerate}
\item Draw regression coefficient $\eta_{i}\mid\mu\sim\mathcal{N}\left(\mu,1\right)$
\item Draw a response variable $y_{i}\mid\bar{z},\eta_{i},Y,\xi\sim\frac{\Phi\left(\eta_{i}^{T}\bar{z}\right)\psi\left(Y\right)}{\mathcal{Z}}$
where $\bar{z}=N^{-1}\sum_{n=1}^{N}z_{n}$ and $\psi\left(Y\right)=\exp\left\{ \sum_{\left\{ p,c\right\} }f(p,c)\right\} $
and $\begin{cases}
\begin{array}{c}
f(0,0)=\xi_{0}\\
f(0,1)=-\infty\\
f(1,0)=\xi_{1}\\
f(1,1)=\xi_{2}\end{array}\end{cases}$
\end{enumerate}
\end{enumerate}
We will employ a data augmentation scheme with auxiliary variables
$a_{i}$ in the probit model such that:

\begin{equation}
y_{i}\sim\begin{cases}
\begin{array}{c}
1,\\
0,\end{array} & \begin{array}{c}
a_{i}\geq0\\
a_{i}<0\end{array}\end{cases}\end{equation}

@@

\begin{equation}
\eta_{i}\sim\mathcal{N}\left(\mu=0,1\right)\end{equation}

@@

The denominator for this distribution is the marginal probability
of the data and cannot be solved in closed form. This is often the
case in evaluating posterior distributions of non-trivial probabilistic
models. We will appeal to one of the common methods for approximating
posterior distributions in the face of intractable normalization factors:
markov chain monte carlo (MCMC). Since in this model it is possible
to sample from the conditional distributions for all variables we
will use the gibbs sampling algorithm to obtain an approximation to
this posterior. In particular, we will derive a Rao-Blackwellized
gibbs sampler for the supervised topic model for general exponential
family response variables and then for our model with a probit response
variable.

@@

To derive the Rao-Blackwellized gibbs sampler in general, we integrate
over the parameters $\theta$ and $\phi_{1:K}$ resulting in the
collapsed joint distribution shown in Equations 5-8.

@@

To derive the gibbs sampler for the general exponential family form
as well as the probit regression model, we derive the individual conditional
probability distributions for all latent variables and parameters.

@@

\begin{equation}
\propto\prod_{i=1}^{I}\left[p\left(y_{m,i}\mid z_{m,1:N},\eta_{i},\xi\right)\right]\left(n_{m,\left(\bullet\right)}^{k,-\left(m,n\right)}+\alpha_{k}\right)\frac{n_{\left(\bullet\right),w_{m,n}}^{k,-\left(m,n\right)}+\beta_{k,w_{m,n}}}{\sum_{v=1}^{V}\left(n_{\left(\bullet\right),w_{m,n}}^{k,-\left(m,n\right)}+\beta_{k,v}\right)}\end{equation}

@@

\section{Results}
%We ran the Gibbs sampler for two days on 500 documents with 20 topics to learn the sLDA model which we then followed by a prediction of ICD-9 codes for 100 documents. The prediction resulted in 38 documents with at least 1 ICD-9 code assignment while some had over 200 assignments.

@@

\section{Conclusion}
%We have proposed a new method for prediction of ICD-9 codes given a medical discharge summary note.  Our method explicitly takes advantage of the ICD-9 hierarchical IS-A relationships as well as the word distributions in each document.  We used Gibbs sampling for a supervised latent dirichlet allocation model to learn a distribution of words over twenty topics and their appropriate ICD-9 codes (response variable). The model was then tested on 100 discharge summaries for which the ICD-9 diagnoses were known.  The amount of ICD-9 codes produced for some discharge summaries was quite high because of the way the model worked with the IS-A hierarchy, highlighting the completeness of ICD-9 coding when using this sLDA model.  Overall, the model had extremely good specificity while the sensitivity was lower than anticipated.  Given our limited set of documents, and very noisy gold standard it would not be accurate to discount the use of sLDA in medical document classification and automatic coding.
%In future work, we plan to significantly expand the set of documents used to train the model (up to 15,000), as well as run several different iterations to learn the optimal topic number.

@@

\begin{small}
\bibliographystyle{plainnat}
\bibliography{refs} 
\end{small}
\end{document}

>>>>

%
\begin{figure}[h]
%tbp] %  figure placement: here, top, bottom, or page
 \centering \includegraphics[width=4in]{model_1} \includegraphics[width=3in]{model_2}
\caption{adapted sLDA model}

\label{fig:example} 
\end{figure}

Given the number of topics, K, the global prior over topic proportions,
$\alpha'$, and the prior over topics, $\gamma$, the generative process
for documents and responses is as follows: 
\begin{enumerate}
\item For each topic:

\begin{enumerate}
\item Draw a distribution over words $\beta_{k}\sim Dir\left(u,\gamma\right)$
\end{enumerate}
\item For each ICD9 Code:

\begin{enumerate}
\item Draw regression coefficient $\eta_{i}\mid\mu,\sigma\sim\mathcal{N}\left(\mu,\sigma\right)$ 
\end{enumerate}
\item Draw a prior over topic proportions $m\mid\alpha'\sim Dir\left(u,\alpha'\right)$ 
\item For each document:

\begin{enumerate}
\item Draw topic proportions $\theta_{d}\mid\alpha\sim Dir\left(m,\alpha\right)$ 
\item For each word:

\begin{enumerate}
\item Draw topic assignment $z_{n,d}\mid\theta_{d}\sim Mult\left(\theta_{d}\right)$ 
\item Draw word $w_{n,d}\mid z_{n,d},\beta_{1:K}\sim Mult\left(\beta_{z_{n}}\right)$ 
\end{enumerate}
\item For each ICD-9 code from the root of the hierarchy and recursively
descending the tree:

\begin{enumerate}
\item Draw a response variable $y_{i}\mid\bar{z},\eta_{i},y_{parent}\sim\Phi(\eta_{i}^{T}\bar{z},y_{parent})$
where $\bar{z}=N^{-1}\sum_{n=1}^{N}z_{n}$ and $\Phi$ refers to a
conditional probit model.
\end{enumerate}
\end{enumerate}
\end{enumerate}
We will employ a data augmentation scheme with auxiliary variables
$a_{i}$ in the probit model where:

\begin{equation}
y_{i}\sim\begin{cases}
\begin{array}{c}
1,\\
0,\end{array} & \begin{array}{c}
a_{i}>0\; and\; y_{parent}=1\\
otherwise\end{array}\end{cases}\end{equation}

@@

@@

The denominator for this distribution is the marginal probability
of the data and cannot be solved in closed form. This is often the
case in evaluating posterior distributions of non-trivial probabilistic
models. We will appeal to one of the common methods for approximating
posterior distributions in the face of intractable normalization factors:
Markov chain Monte Carlo (MCMC) sampling. Since in this model it is
possible to sample from the conditional distributions for all variables
we will use the Gibbs sampling algorithm to obtain an approximation
to this posterior. In particular, we will derive a collapsed Gibbs
sampler for the supervised topic model.

@@

To derive the Gibbs sampler in general, we integrate over the parameters
$\theta$ and $\phi_{1:K}$ resulting in the collapsed joint distribution
shown in Equations 5-8.

@@

To derive the Gibbs sampler we evaluate the individual conditional
probability distributions for all unobserved variables.

@@

\begin{equation}
\propto\prod_{i=1}^{I}\left[p\left(y_{m,i}\mid z_{m,1:N},\eta_{i},\xi\right)\right]\left(n_{m,\left(\bullet\right)}^{k,-\left(m,n\right)}+\alpha_{k}\right)\frac{n_{\left(\cdot\right),w_{m,n}}^{k,-\left(m,n\right)}+\beta_{k,w_{m,n}}}{\sum_{v=1}^{V}\left(n_{\left(\cdot\right),w_{m,n}}^{k,-\left(m,n\right)}+\beta_{k,v}\right)}\end{equation}

@@

\section{Results}

%We ran the Gibbs sampler for two days on 500 documents with 20 topics to learn the sLDA model which we then followed by a prediction of ICD-9 codes for 100 documents. The prediction resulted in 38 documents with at least 1 ICD-9 code assignment while some had over 200 assignments.

@@

\section{Conclusion}

%We have proposed a new method for prediction of ICD-9 codes given a medical discharge summary note.  Our method explicitly takes advantage of the ICD-9 hierarchical IS-A relationships as well as the word distributions in each document.  We used Gibbs sampling for a supervised latent dirichlet allocation model to learn a distribution of words over twenty topics and their appropriate ICD-9 codes (response variable). The model was then tested on 100 discharge summaries for which the ICD-9 diagnoses were known.  The amount of ICD-9 codes produced for some discharge summaries was quite high because of the way the model worked with the IS-A hierarchy, highlighting the completeness of ICD-9 coding when using this sLDA model.  Overall, the model had extremely good specificity while the sensitivity was lower than anticipated.  Given our limited set of documents, and very noisy gold standard it would not be accurate to discount the use of sLDA in medical document classification and automatic coding.
%In future work, we plan to significantly expand the set of documents used to train the model (up to 15,000), as well as run several different iterations to learn the optimal topic number.

@@

\begin{small} \bibliographystyle{plainnat} \bibliographystyle{plainnat}
\bibliography{refs}
 \end{small} 
\end{document}
